---
title: "How to integrate a platemap which maps out experiments to name columns automatically and group results."
author: "Mark Cheng"
date: "17/03/2024"
output:
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
#Unorthodox calling in a vignette: did not use standard output of html_vignette (which is simpler but more space efficient), but rather traditional .rmd html_document format.
#This is because this is not uploaded to CRAN. see more details here (https://bookdown.org/yihui/rmarkdown/r-package-vignette.html).

---
## Setting up
load the library/package from github.
```{r setup}
if(!require("devtools"))install.packages("devtools",repos="http://cran.us.r-project.org")
devtools::install_github("TKMarkCheng/NormaliseForIC50",dependencies = TRUE,force = TRUE)
```

```{r}
NormaliseForIC50::generate_plate_map("../Validation/",output_plateMap_file = "test.xlsx")
```


### Loading sample, defining output file, and plate rotation (if any)
It is very important for you to change the following file paths and settings to fit your data. (Where are they stored? Where and what do you what the output excel file to be called? Are your plates rotated?)
```{r setting up files}


input_file <- "Validation/2022-09-04 reads/example1.xlsx" #CHANGE to example file in your directory for troubleshooting
input_directory = "Validation/2022-09-04 reads/" #CHANGE this to directory of input .xlsx data files
output_file <- "Validation/validation_output/test.xlsx" # CHANGE directory and name for yourself
```

It is also important that you communicate how your plates are setup, and they should be consistent for each folder that you use this script on. i.e. don't have horizontal plates and vertical plates mixed in the same folder, and positive and negative controls should be in consistent columns in each folder this script is used on.
```{r plate setup}
rotate_by <- 0 # clockwise degrees, multiples of 90, useful when doing 12 fold dilutions.
control_neg_column <- c(1) # default negative control column is on first column from the left
control_pos_column <- c(2) # default positive control column is on second column from the left
```
## Getting the filenames from our directory
What are the names of all of our input files?
```{r get all the filepaths in the directory}
neut_raw_files <- Sys.glob(paste0(
  paste0(input_directory,"*.xlsx")) # find all .xlsx file in input directory
)
neut_raw_files
```
## Example case: Plate 1
An inital view of plate 1
```{r example}
plate1 <- NormaliseForIC50::read_promega_plate_excel(input_promega_excel_file_path = input_file)
plate1
```
The Name assigned to each column can be any string of characters, as long as there's no excessively weird symbols.

```{r test run}
NormaliseForIC50::final_func(neut_raw_files[1],control_neg_column = c(1),control_pos_column = c(2), rotation_deg_needed = rotate_by)
```
## Processing all the files in the directory

```{r}
#create xlsx and populate df into separate sheets. CHANGE rotation_deg_needed if the plate was ROTATED when setting up.
library(openxlsx)
wb<-createWorkbook()
for (neut_raw_file in neut_raw_files){
  neut_file <- basename(neut_raw_file) #remove file path
  neut_file = tools::file_path_sans_ext(neut_file) #remove extension
  addWorksheet(wb,sheetName=neut_file)
  writeData(wb, sheet = neut_file,
            x = NormaliseForIC50::final_func(
              neut_raw_file,
              control_neg_column = control_neg_column,
              control_pos_column = control_pos_column,
              rotation_deg_needed = rotate_by))
  print(paste0("normalised data written to:",neut_file))
}
saveWorkbook(wb,file=output_file,overwrite = TRUE)
print(paste0("saved collated sheets to excel file:",output_file))
```
