---
title: "How to integrate a platemap which maps out experiments to name columns automatically and group results."
author: "Mark Cheng"
date: "17/03/2024"
output:
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
#Unorthodox calling in a vignette: did not use standard output of html_vignette (which is simpler but more space efficient), but rather traditional .rmd html_document format.
#This is because this is not uploaded to CRAN. see more details here (https://bookdown.org/yihui/rmarkdown/r-package-vignette.html).

---
## Setting up
load the library/package from github.
```{r setup}
# if(!require("devtools"))install.packages("devtools",repos="http://cran.us.r-project.org")
devtools::install_github("TKMarkCheng/NormaliseForIC50",dependencies = TRUE,force = TRUE)
```


### Loading sample, defining output file.
```{r setting up files}
input_file <- "../Validation/Concentration_mode/2024-03-16/Inhib1_Inhib2_Vir1Vir2.xlsx" #CHANGE to example file in your directory for troubleshooting
input_directory = "../Validation/" #CHANGE this to directory of input .xlsx data files
output_generated_plateMap_file <- "../Validation/validation_output/example_generated_platemap.xlsx" # CHANGE directory and name for yourself

```

## Generate platemap based on the files in a directory. 
This function fixes missing columns, and also finds the promega_plate_path by searching the Plate_Name you have given. You need to make sure your plate names are unique.
```{r}
bruh <- NormaliseForIC50::generate_plate_map(input_directory,output_mode = "return")
bruh
```
you can choose to write to an excel file and manually add values.
```{r}
#NormaliseForIC50::generate_plate_map(input_directory,output_mode = "write",output_plateMap_file = output_generated_plateMap_file)
```

```{r}
platemap <- readxl::read_excel(path="../Validation/validation_output/example_generated_platemap_manual_changes.xlsx")
platemap
```
Or you can feed in a pre-made excel file.
```{r}
library(dplyr)

platemap_sera <- platemap %>% filter(Group=="inhibition")
platemap_sera
```


```{r sample of 1 plate}
plate1 <- unique(platemap_sera$Plate_Name)[1]
plate1_map <- platemap_sera %>% filter(Plate_Name == plate1)

plateMap_map_onto_promega_read <- function(plateMap,promega_read_values){
  #plateMap = platemap section for 1 plate
  #proemga_read_values = read in via NormaliseForIC50::read_promgea_plate_excel. This is defaulted to be pulled from the plateMap's column$promega_plate_path
  plate_map <- plateMap %>% mutate(column=as.numeric(gsub("\\D","",Well))) # create a column reference
  
  processed_read_values <- promega_read_values
  # change the column headers to <Individual_condition>_<Virus>
  colnames(processed_read_values) <- paste0(plate_map$Individual_condition[match(1:ncol(processed_read_values),plate_map$column)],
                                  "_",
                                  plate_map$Virus[match(1:ncol(processed_read_values),plate_map$column)])
  
  # adding the positive and negative control (virus specific manner)
  for (negative_control_column in unique(plate_map$Negative_Control_Column)){
  negative_control <- plate_map %>% filter(Negative_Control_Column == negative_control_column) %>% select(Virus) %>% unique()
  negative_control_virus <- negative_control[[1,1]]
  colnames(processed_read_values)[negative_control_column] <- paste0("Neg_",negative_control_virus)
  }
  for (positive_control_column in unique(plate_map$Positive_Control_Column)){
  positive_control <- plate_map %>% filter(Positive_Control_Column == positive_control_column) %>% select(Virus) %>% unique()
  positive_control_virus <- positive_control[[1,1]]
  colnames(processed_read_values)[positive_control_column] <- paste0("Pos_",positive_control_virus)
  }
  return(list(plate_map, processed_read_values))
}

plate1_map <- plateMap_map_onto_promega_read(plate1_map,NormaliseForIC50::read_promega_plate_excel(input_promega_excel_file_path = plate1_map$promega_plate_path[1]))[[1]]
plate1_values <- plateMap_map_onto_promega_read(plate1_map,NormaliseForIC50::read_promega_plate_excel(input_promega_excel_file_path = plate1_map$promega_plate_path[1]))[[2]]

plate1_values

adding_technical_replicates <- function(mapped_plate){
  custom_suffix="_techrep"
  # Get column names and their frequencies
  col_names <- names(mapped_plate)
  col_freq <- table(col_names)
  # Identify duplicated column names
  duplicated_cols <- names(col_freq)[col_freq > 1]
  
  # Iterate over duplicated column names and add custom suffix
  for (col in duplicated_cols) {
    indices <- which(col_names == col)
    for (i in seq_along(indices)) {
      col_names[indices[i]] <- paste(col, paste0(custom_suffix, i), sep = "")
      }
  }
  # Update the dataframe with modified column names
  names(mapped_plate) <- col_names
  return(mapped_plate)
}

plate1_values <- adding_technical_replicates(plate1_values)
plate1_values
```
```{r continue example of previous plate}
# now treat each unique virus suffix as their own set, and normalise.
normalise_plate_using_plateMap <- function(plate_map, plate_values){
  df_list <- list()
  for (virus in unique(plate_map$Virus)){
    plate_map_virus_specific <- plate_map %>% filter(Virus==virus)
    neg_control_columns <- unique(plate_map_virus_specific$Negative_Control_Column)
    pos_control_columns <- unique(plate_map_virus_specific$Positive_Control_Column)
    plate_map_virus_specific_normalise_columns <- c(neg_control_columns,
                                                     pos_control_columns,
                                                     plate_map_virus_specific$column) # powerful because this also reorders them in order of negative, positive, experiment columns
    bruh <- NormaliseForIC50::normalise(df = plate_values[plate_map_virus_specific_normalise_columns],
                                        control_neg_column = seq(length(neg_control_columns)),
                                        control_pos_column = seq(length(neg_control_columns)+1,length(neg_control_columns)+ length(pos_control_columns)))
    df_list[[virus]] <- bruh 
  }
  plate_normalised <- do.call("cbind",df_list)
  return(plate_normalised)
}
plate1_normalised <- normalise_plate_using_plateMap(plate1_map,plate1_values)
plate1_normalised
```

validated up to this point. Dilution factor also completed below but needs further integration.

```{r}
# create a new far-left column called dilution
convert_dilution_factor <- function(dilution_factor_string="1 in 3"){
  numerator <- as.numeric(stringr::str_extract(dilution_factor_string,"^[0-9]+(?:\\.[0-9]+)?")) # first integer or decimal numeric
  denominator <- as.numeric(stringr::str_extract(dilution_factor_string,"[0-9]+(?:\\.[0-9]+)?$")) # last integer or decimal numeric
  dilution_factor <- denominator/numerator
  return(dilution_factor)
}

generate_dilution_series <- function(Starting_Dilution_or_concentration = 20, num_steps = 8, dilution_factor = "1 in 3", titration_mode = c("dilution","concentration")){
 dilution_factor <- convert_dilution_factor(dilution_factor)
 dilution_series <- numeric(num_steps)
 dilution_series[1] <- Starting_Dilution_or_concentration
 for (i in 2:num_steps){
  dilution_series[i] <- switch(titration_mode,
                               "dilution" = dilution_series[i-1] * dilution_factor, 
                               "concentration" = dilution_series[i-1] / dilution_factor)
 }
 return(dilution_series)
}

num_steps <- nrow(plate1_values)

dilution_df <- plate1_map%>% select(column,Negative_Control_Column,Positive_Control_Column,Virus,dilution_or_concentration,Starting_Dilution_or_concentration,dilution_series) %>%
  group_by(Negative_Control_Column,Positive_Control_Column,Virus,dilution_or_concentration,Starting_Dilution_or_concentration,dilution_series) %>%
  summarise(column=paste(column,collapse=","))%>%
  ungroup()

dilution_df <- dilution_df %>% rowwise() %>% mutate(
  dilution_serie = paste(generate_dilution_series(titration_mode = dilution_or_concentration,num_steps = 8,
                                             Starting_Dilution_or_concentration = Starting_Dilution_or_concentration,
                                             dilution_factor = dilution_series)
                         ,collapse = ",")
  )
dilution_df

NormaliseForIC50::normalise(plate1_values,control_neg_column = dilution_df$Negative_Control_Column,control_pos_column = dilution_df$Positive_Control_Column)

```



It is also important that you communicate how your plates are setup, and they should be consistent for each folder that you use this script on. i.e. don't have horizontal plates and vertical plates mixed in the same folder, and positive and negative controls should be in consistent columns in each folder this script is used on.
```{r plate setup}
rotate_by <- 0 # clockwise degrees, multiples of 90, useful when doing 12 fold dilutions.
control_neg_column <- c(1) # default negative control column is on first column from the left
control_pos_column <- c(2) # default positive control column is on second column from the left
```
## Getting the filenames from our directory
What are the names of all of our input files?
```{r get all the filepaths in the directory}
neut_raw_files <- Sys.glob(paste0(
  paste0(input_directory,"*.xlsx")) # find all .xlsx file in input directory
)
neut_raw_files
```
## Example case: Plate 1
An inital view of plate 1
```{r example}
plate1 <- NormaliseForIC50::read_promega_plate_excel(input_promega_excel_file_path = input_file)
plate1
```
The Name assigned to each column can be any string of characters, as long as there's no excessively weird symbols.

```{r test run}
NormaliseForIC50::final_func(neut_raw_files[1],control_neg_column = c(1),control_pos_column = c(2), rotation_deg_needed = rotate_by)
```
## Processing all the files in the directory

```{r}
#create xlsx and populate df into separate sheets. CHANGE rotation_deg_needed if the plate was ROTATED when setting up.
library(openxlsx)
wb<-createWorkbook()
for (neut_raw_file in neut_raw_files){
  neut_file <- basename(neut_raw_file) #remove file path
  neut_file = tools::file_path_sans_ext(neut_file) #remove extension
  addWorksheet(wb,sheetName=neut_file)
  writeData(wb, sheet = neut_file,
            x = NormaliseForIC50::final_func(
              neut_raw_file,
              control_neg_column = control_neg_column,
              control_pos_column = control_pos_column,
              rotation_deg_needed = rotate_by))
  print(paste0("normalised data written to:",neut_file))
}
saveWorkbook(wb,file=output_file,overwrite = TRUE)
print(paste0("saved collated sheets to excel file:",output_file))
```
