---
title: "How to integrate a platemap which maps out experiments to name columns automatically and group results."
author: "Mark Cheng"
date: "17/03/2024"
output:
  html_document: 
    theme: cerulean
    toc: yes
    toc_float: yes
#Unorthodox calling in a vignette: did not use standard output of html_vignette (which is simpler but more space efficient), but rather traditional .rmd html_document format.
#This is because this is not uploaded to CRAN. see more details here (https://bookdown.org/yihui/rmarkdown/r-package-vignette.html).

---
## Setting up
load the library/package from github.
```{r setup}
if(!require("devtools"))install.packages("devtools",repos="http://cran.us.r-project.org")
devtools::install_github("TKMarkCheng/NormaliseForIC50",dependencies = TRUE,force = TRUE)
```

### Loading sample, defining output file.
```{r setting up files}
input_directory = paste0(here::here("Validation/"),"/") #CHANGE this to directory of input .xlsx data files
output_generated_plateMap_file <- here::here("Validation/validation_output/example_generated_platemap.xlsx") # CHANGE directory and name for yourself
```

## Generate platemap based on the files in a directory. 
This function fixes missing columns, and also finds the promega_plate_path by searching the Plate_Name you have given. You need to make sure your plate names are unique.
```{r}
library(dplyr)
bruh <- NormaliseForIC50::generate_plate_map(input_directory,output_mode = "return")
bruh
```
you can choose to write to an excel file and manually add values. Or you can feed in a pre-made excel file.

```{r}
#NormaliseForIC50::generate_plate_map(input_directory,output_mode = "write",output_plateMap_file = output_generated_plateMap_file)
## manual correction/filling
platemap <- readxl::read_excel(path="../Validation/validation_output/example_generated_platemap_manual_changes.xlsx")
platemap
```

```{r}
options(scipen = 999)
df_list <- list()
for (each_plate_name in unique(platemap$Plate_Name)){
  print(each_plate_name)
  individual_platemap <- platemap %>% dplyr::filter(Plate_Name == each_plate_name)
  plate_normalised <- normalise_plate_using_plateMap(individual_platemap)
  #plate_normalised <- cbind(plate_name=each_plate_name,plate_normalised)
  print(plate_normalised)
  df_list[[each_plate_name]] <- plate_normalised
}
merge_cols = c("Group","dilution_or_concentration","dilution_serie")
plates_merged <- Reduce(function (x,y) merge(x,y,by=merge_cols, all = TRUE),df_list) %>% dplyr::arrange(across(all_of(merge_cols)))
plates_merged

```
```{r}
# if you want to pick by specific conditions or viruses
filter_by_condition = NULL
filter_by_virus = c("Virus1","VIrus2")

filter_merged <- function(plates_merged,condition=NULL,virus=NULL){
  experiment_columns <- data.frame(colnames=colnames(plates_merged)) %>% 
    subset(.,!(colnames %in% merge_cols)) %>%
    tidyr::separate(colnames, into=c("individual_condition","Virus","technical_replicate"),"_",remove=FALSE)
  
  filtered_experiment_columns <- experiment_columns %>%
    filter(if(!is.null(condition)) (individual_condition %in% condition) else TRUE) %>%
    filter(if(!is.null(virus)) (Virus %in% virus) else TRUE)
  
  filtered_plates_merged <- plates_merged %>% 
    dplyr::select(merge_cols,filtered_experiment_columns$colnames) %>%
    dplyr::filter(!if_all(-merge_cols,is.na))
  
  #remove technical replicate in colnames
  colnames(filtered_plates_merged) <- sub("_techrep[0-9]+$", "", colnames(filtered_plates_merged)) 
  
  return (filtered_plates_merged)
}
plates_merged_filter <- filter_merged(plates_merged,condition=filter_by_condition,virus = filter_by_virus)
plates_merged_filter

write.xlsx(plates_merged_filter,
           here::here(paste0(
             "Validation/validation_output/PlatesMerged",
             "_",(if (!is.null(filter_by_condition)) paste(filter_by_condition,collapse='-') else 'AllCondition'),
             "_",(if (!is.null(filter_by_virus)) paste(filter_by_virus,collapse='-') else 'AllVirus'),
             ".xlsx"))
           )
```

Alternatively, you can group by the virus or individual condition and generate sheets per virus or individual condition.
```{r}
#groupby viruses
output_file_groupby_virus <- here::here("Validation/validation_output/PlatesMerged_Grouped_by_virus.xlsx")

library(openxlsx)
wb<-createWorkbook()
virues <- unique(platemap$Virus)
for (virus in virues){
  plates_merged_filter <- filter_merged(plates_merged,virus = virus)
  sheetName=virus
  addWorksheet(wb,sheetName=sheetName)
  writeData(wb, sheet = sheetName,
            x=plates_merged_filter)
  print(glue::glue("data for virus:{virus} calculated, grouped and written."))
  print(plates_merged_filter)
}
saveWorkbook(wb,file=output_file_groupby_virus,overwrite = TRUE)
```


```{r}
#groupby condition
output_file_groupby_condition <- here::here("Validation/validation_output/PlatesMerged_Grouped_by_condition.xlsx")

library(openxlsx)
wb<-createWorkbook()
conditions <- unique(platemap$Individual_condition)
for (condition in conditions){
  plates_merged_filter <- filter_merged(plates_merged,condition = condition)
  sheetName=condition
  addWorksheet(wb,sheetName=sheetName)
  writeData(wb, sheet = sheetName,
            x=plates_merged_filter)
  print(plates_merged_filter)
  print(glue::glue("data for condition:{condition} calculated, grouped and written."))
}
saveWorkbook(wb,file=output_file_groupby_condition,overwrite = TRUE)
```
